---
title: "Multivariate Analysis of Contaminants in Farmed and Wild Salmon"
author: "Kaylyn Vo"
date: "Fall, 2020"
output:
  html_document:
    toc: true
    toc_depth: 5
    theme: united
---



# 1. Libraries

```{r}
library('dplyr')
library('car')
library('onewaytests')
```

# 2. Data

```{r}
feed = read.table('R_FEED.csv',header = TRUE,sep=',',stringsAsFactors = FALSE)
farmed = read.table('R_FARMED.csv',header = TRUE,sep=',',stringsAsFactors = FALSE)
wild = read.table('R_WILD.csv',header = TRUE,sep=',',stringsAsFactors = FALSE)
supermarket = read.table('R_SUPERMARKET.csv',header = TRUE,sep=',',stringsAsFactors = FALSE)
```

```{r,echo=FALSE}

col_names = c("Dieldrin","Total.Toxaphene.Equivalvents","DIOXIN.TEQ.1","TOTAL.PCBs","TOTAL.OMEGA.6.FATTY.ACIDS","TOTAL.OMEGA.3.FATTY.ACIDS","PERCENT.LIPID")

names <- c('Dieldrin','Toxaphene','Dioxin','PCBs','Omega-6','Omega-3',"%Lipid")

######### OBSERVATION 153 IN THE FARMED FILE DID NOT MEET QA/QS, SO WE DROP THIS OBSERVATION 
farmed =farmed[-153,]

fish_type <- matrix(rep(c('WILD','FARMED','FARMED'),c(45,48,152)))

c_salmon <- cbind(fish_type,rbind(wild[,c('LOCATION',col_names)],supermarket[,c('LOCATION',col_names)],farmed[,c('LOCATION',col_names)]))
```


# 3. Helper Functions
```{r}
calc_cor <- function(X)
{
  corr <- cor(X)
  colnames(corr) <- names
  rownames(corr) <- names
  corr <- round(corr,3)
  return(corr)
}
##### Function to calculate variance

calc_var <- function(X)
{
  S <- var(X)
  colnames(S) <- names
  rownames(S) <- names
  S <- round(S,3)
  return(S)
}

### Function to calculate Normality plot
plot_norm <- function(X,toxin_name)
{
  title_ = paste('Normal Q-Q Plot for',toxin_name, sep=" ")
  qqnorm(X,main = title_)
  qqline(X)
}

### Function to calculate the chi-square plot
plot_chisq <- function(X)
{
  x_bar <- as.matrix(colMeans(X))
  S <- var(X)
  n <- nrow(X)
  p <- ncol(X)
  # Then, calculate all Mahalanobis distances
  d <- sapply(1:n, function(k) (t(as.matrix(X[k,]) - x_bar) %*% solve(S) %*%(X[k,] - x_bar)))
  q <- qchisq((1:n-0.5)/n,p)
  qqplot(q,d,xlab = "Chi-square Quantiles", ylab = "Sample Statistical Distances", main = "Chi-square Probability Plot")
  lines(q,q)
  
}

### Function to tranform data

X_transform <- function(X,tranform_vars,o_vars)
{
  t_X <- cbind(log(X[,tranform_vars]),X[,o_vars])
  return(t_X)
}


calc_CIs <- function(S,d,c,p)
{
  mean_CIs = matrix(0, nrow = p, ncol = 2)
  colnames(mean_CIs) <- c('Lower CI','Upper CI')
  rownames(mean_CIs) <- names
  
  for (i in 1:p)
  {
    mean_CIs[i,1] = d[i] - sqrt(c) * sqrt(S[i,i])
    mean_CIs[i,2] = d[i] + sqrt(c) * sqrt(S[i,i])
  }
  
  mean_CIs <- round(mean_CIs,3)
  return(mean_CIs)
  
}
####### Function to combine data
combine_df <- function(farmed_filt,supermarket_filt)
{
  df = as.matrix(sapply(rbind(farmed[farmed_filt,col_names],supermarket[supermarket_filt,col_names]),as.numeric))
  return(df)
}
```

```{r,echo=FALSE}
f = c_salmon$fish_type=='FARMED'

f_X = as.matrix(sapply(c_salmon[f,col_names],as.numeric))
w_X = as.matrix(sapply(c_salmon[!f,col_names],as.numeric))

t_vars = c("Dieldrin","Total.Toxaphene.Equivalvents","DIOXIN.TEQ.1","TOTAL.PCBs","TOTAL.OMEGA.6.FATTY.ACIDS")
o_vars = c("TOTAL.OMEGA.3.FATTY.ACIDS","PERCENT.LIPID")


t_farmed <- X_transform(f_X,t_vars,o_vars)
t_wild <- X_transform(w_X,t_vars,o_vars)
```

# 4. Principal Component Analysis

```{r}
salmon = as.matrix(sapply(c_salmon[,col_names],as.numeric))
colnames(salmon) <- names
pc_salmon <- prcomp(salmon,scale = TRUE)
```

## 4.A. Cumulative Total Sample Variance

```{r}
(cumvar <- cumsum(pc_salmon$sdev^2)/sum(pc_salmon$sdev^2))
```

## 4.B. Sample Proportion of Total Variance

```{r}
propvar <- pc_salmon$sdev^2/sum(pc_salmon$sdev^2)
```

# 5. Comparisons of Contaminants between Wild and Farmed Salmon

## 5.A. Normality Assumption 

### 5.A.1 Farmed Salmon

```{r, fig.show="hold",out.width="90%"}
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
plot_norm(t_farmed[,'Dieldrin'],'Log Dieldrin')
plot_norm(t_farmed[,'Total.Toxaphene.Equivalvents'],'Log Toxaphene')
plot_norm(t_farmed[,'DIOXIN.TEQ.1'],'Log Dioxin')
plot_norm(t_farmed[,'TOTAL.PCBs'],'Log PCBs')
plot_norm(t_farmed[,'TOTAL.OMEGA.6.FATTY.ACIDS'],'Log Omega-6')
plot_norm(t_farmed[,'TOTAL.OMEGA.3.FATTY.ACIDS'],'Omega-3')
plot_norm(t_farmed[,'PERCENT.LIPID'],'Lipid Pct.')
plot_chisq(t_farmed)

```

### 5.A.2 Wild Salmon
```{r, fig.show="hold",out.width="90%"}
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
plot_norm(t_wild[,'Dieldrin'],'Log Dieldrin')
plot_norm(t_wild[,'Total.Toxaphene.Equivalvents'],'Log Toxaphene')
plot_norm(t_wild[,'DIOXIN.TEQ.1'],'Log Dioxin')
plot_norm(t_wild[,'TOTAL.PCBs'],'Log PCBs')
plot_norm(t_wild[,'TOTAL.OMEGA.6.FATTY.ACIDS'],'Log Omega-6')
plot_norm(t_wild[,'TOTAL.OMEGA.3.FATTY.ACIDS'],'Omega-3')
plot_norm(t_wild[,'PERCENT.LIPID'],'Lipid Pct.')
plot_chisq(t_wild)
```


## 5.B. Sample Variance

### 5.B.1. Farmed Salmon

```{r,echo=FALSE}

calc_var(t_farmed)

```

### 5.B.2. Wild Salmon 
```{r,echo=FALSE}

calc_var(t_wild)

```

## 5.C. Test Statistics

```{r,echo=FALSE}
n1 = nrow(t_wild)      
n2 = nrow(t_farmed)     
p=ncol(t_farmed) 

farmed_variance <- calc_var(t_farmed)
wild_variance <- calc_var(t_wild)

xbar_farmed = as.matrix(colMeans(t_farmed))
xbar_wild = as.matrix(colMeans(t_wild))

mean_diff = xbar_farmed - xbar_wild

combine_Var <- (1/n1)*wild_variance + (1/n2)*farmed_variance

T_squared <- t(mean_diff) %*% solve(combine_Var) %*% (mean_diff)
c = qchisq(0.95,p)
T_squared

```
## 5.D. Confidence Intervals

```{r,echo=FALSE}
calc_CIs(combine_Var,mean_diff,c,p)
```

# 6. Comparisons of Contaminants between Europe and North America 

```{r,echo=FALSE}
Europe <- (farmed$LOCATION =="Scotland") | (farmed$LOCATION== 'Norway') | (farmed$LOCATION== "Faroe Islands")
North_America <- (farmed$LOCATION == "Western Canada") | (farmed$LOCATION=='Washington') |
  (farmed$LOCATION =='Maine') | (farmed$LOCATION =="Eastern Canada") 
South_America <- (farmed$LOCATION == 'Chile')  

supermarket$LOCATION <- trimws(supermarket$LOCATION)

d = (supermarket$LOCATION == "Norway" | supermarket$LOCATION == "Scotland"| supermarket$LOCATION == 'Germany'| supermarket$LOCATION == "Iceland")

f = (supermarket$LOCATION == 'Maine' | supermarket$LOCATION == 'Eastern Canada'|supermarket$LOCATION == 'N.E. Atlantic'|supermarket$LOCATION == "American Atlantic"| 
       supermarket$LOCATION =="Atlantic" |supermarket$LOCATION =="Western Canada"|supermarket$LOCATION == "Washington" | supermarket$LOCATION =='BC')
g = (supermarket$LOCATION == "Chile")  


## Check if we have the right number of observations
#sum(d) + sum(Europe) # 81
#sum(f) + sum(North_America) # 73
#sum(g) + sum(South_America) # 35

### Europe
df_E = combine_df(Europe,d)

### North America

df_NAM = combine_df(North_America,f)

### South America

df_SAM = combine_df(South_America,g)

t_Europe<- X_transform(df_E,t_vars,o_vars)
t_NAM <- X_transform(df_NAM,t_vars,o_vars)
t_SAM <- X_transform(df_SAM,t_vars,o_vars)

```

## 6.A. Normality Assumption 

### 6.A.1 Europe


```{r, fig.show="hold",out.width="90%"}
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
plot_norm(t_Europe[,'Dieldrin'],'Log Dieldrin')
plot_norm(t_Europe[,'Total.Toxaphene.Equivalvents'],'Log Toxaphene')
plot_norm(t_Europe[,'DIOXIN.TEQ.1'],'Log Dioxin')
plot_norm(t_Europe[,'TOTAL.PCBs'],'Log PCBs')
plot_norm(t_Europe[,'TOTAL.OMEGA.6.FATTY.ACIDS'],'Log Omega-6')
plot_norm(t_Europe[,'TOTAL.OMEGA.3.FATTY.ACIDS'],' Omega-3')
plot_norm(t_Europe[,'PERCENT.LIPID'],'Lipid Pct.')
plot_chisq(t_Europe)
```

### 6.A.2 North America

```{r,fig.show="hold",out.width="90%"}
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
plot_norm(t_NAM[,'Dieldrin'],'Log Dieldrin')
plot_norm(t_NAM[,'Total.Toxaphene.Equivalvents'],'Log Toxaphene')
plot_norm(t_NAM[,'DIOXIN.TEQ.1'],'Log Dioxin')
plot_norm(t_NAM[,'TOTAL.PCBs'],'Log PCBs')
plot_norm(t_NAM[,'TOTAL.OMEGA.6.FATTY.ACIDS'],'Log Omega-6')
plot_norm(t_NAM[,'TOTAL.OMEGA.3.FATTY.ACIDS'],'Omega-3')
plot_norm(t_NAM[,'PERCENT.LIPID'],'Lipid Pct.')
plot_chisq(t_NAM)
```

### 6.A.3 South America
```{r,fig.show="hold",out.width="90%"}
par(mar = c(5.1, 4.1, 4.1, 2.1))
par(mfrow = c(1, 2))
plot_norm(t_SAM[,'Dieldrin'],'Log Dieldrin')
plot_norm(t_SAM[,'Total.Toxaphene.Equivalvents'],'Log Toxaphene')
plot_norm(t_SAM[,'DIOXIN.TEQ.1'],'Log Dioxin')
plot_norm(t_SAM[,'TOTAL.PCBs'],'Log PCBs')
plot_norm(t_SAM[,'TOTAL.OMEGA.6.FATTY.ACIDS'],'Log Omega-6')
plot_norm(t_SAM[,'TOTAL.OMEGA.3.FATTY.ACIDS'],'Omega-3')
plot_norm(t_SAM[,'PERCENT.LIPID'],'Lipid Pct.')
plot_chisq(t_SAM)

```

## 6.B. Sample Variance

### 6.B.1 Europe
```{r,echo=FALSE}
(S_eu <- calc_var(t_Europe))

```


### 6.B.2 North America
```{r,echo=FALSE}
(S_nam <- calc_var(t_NAM))
```


### 6.B.3 South America
```{r,echo=FALSE}
(S_sam <- calc_var(t_SAM))
```

## 6.C. Test Statistics
```{r,echo=FALSE}
xbar_eu = as.matrix(colMeans(t_Europe))
xbar_nam = as.matrix(colMeans(t_NAM))

N_eu <- nrow(t_Europe)
N_nam <- nrow(t_NAM)

mean_diff <- xbar_eu - xbar_nam

c_S <- ((1/N_eu) * S_eu) + ((1/N_nam)*S_nam)

(T_squared <- t(mean_diff) %*% solve(c_S) %*% (mean_diff))

c = qchisq(0.95,p)
p = 7
```

## 6.D Confidence Intervals

```{r,echo=FALSE}
calc_CIs(c_S,mean_diff,c,p)
```